Arriaga meets Kitagawa. life expectancy decompositions including population subgroups

### 2024-01-24 manuscript/manuscript.Rmd
updates

# Sample for structured abstract
abstract: |
 **Background**: An Arriaga (1984) decomposition allows us to decompose differences in life expectancy into contributions due to mortality rate differences in each age. A Kitawaga (1955) decomposition allows us to decompose differences in a weighted mean into effects from differences in structure and differences in the weighted value.

@@

  %% \theoremstyle{thmstylethree}%
  \theoremstyle{definition}
  \newtheorem{definition}{Definition}%

output: rticles::springer_article
---

@@

Arriaga decompositions can be implemented in a number of relevant ways, per Riffe et. al. (2024), from which we here consider only symmetrical Arriaga decompositions. This implies following the original formulas twice, once in each direction, and averaging the sign-adjusted results. For this, we use the following lifetable columns:

@@

\end{itemize}
```
We also use the superscript $s$ to index subpopulations comprising the total population, and the superscript $t$ to index time points, or some other way of differentiating the total population. The index $e^{s,t}_0$ reads as "life expectancy at birth for subpopulation $s$ at time $t$", e.g. $e^{low, 1990}_0$ could be the life expectancy in 1990 of a low education group.

Then one can write Arriaga's standard decomposition method like as per eq~\eqref{eq_arriaga1}. To keep things simple, we assume a lifetable radix of 1, meaning $\ell_0 = 1$.

$$
\begin{aligned}
{}_n\Delta_x &= e_0^{s,2} - e_0^{s,1} \\
 &= \ell_x^{s,1} \cdot \left( \frac{{}_nL_x^{s,2}}{\ell_x^{s,2}} - \frac{{}_nL_x^{s,1}}{\ell_x^{s,1}}\right)+T_{x+n}^{s,2}\cdot\left(\frac{\ell_x^{s,1}}{\ell_x^{s,2}} - \frac{\ell_{x+n}^{s,1}}{\ell_{x+n}^{s,2}}\right)
\end{aligned}
$$

@@

Authors reporting data from electrophoretic gels and blots should supply the full unprocessed scans for key as part of their Supplementary information. This may be requested by the editorial team/s if it is missing.

@@

::: {.flushleft latex="true"}
Editorial Policies for:

\noindent Springer journals and proceedings: <https://www.springer.com/gp/editorial-policies>

\noindent Nature Portfolio journals: <https://www.nature.com/nature-research/editorial-policies>

\noindent \textit{Scientific Reports}: <https://www.nature.com/srep/journal-policies/editorial-policies>

\noindent BMC journals: <https://www.biomedcentral.com/getpublished/editorial-policies>
:::

>>>>

# Sample for structured abstract
abstract: |
 **Background**: An Arriaga (1984) decomposition allows us to decompose differences in life expectancy into contributions due to mortality rate differences in each age. A Kitawaga (1955) decomposition allows us to decompose differences in a weighted mean into an effect from differences in structure and effects from differences in each element of the weighted value.

@@

  %% \theoremstyle{thmstylethree}%
  \theoremstyle{definition}
  \newtheorem{definition}{Definition}%
  \usepackage{amsmath}
output: rticles::springer_article
---

@@

Arriaga decompositions can be implemented in a number of relevant ways, per Riffe et. al. (2024), from which we here consider only _symmetrical_ Arriaga decomposition. This implies following the original formulas twice, once in each direction, and averaging the sign-adjusted results. For this, we use the following lifetable columns:

@@

\end{itemize}
```
We also use the superscript $s$ to index subpopulations comprising the total population, and the superscript $t$ to index time points, or some other way of differentiating the total population. The index $e^{s,t}_0$ reads as "life expectancy at birth for subpopulation $s$ at time $t$", e.g. $e^{low, 1990}_0$ could be the life expectancy in 1990 of a low education group. The first stage of our exercise is to decompose subgroup-specific differences in life expectancy (hence the $^s$ on everything):

\begin{align}
{}\Delta^s &= e_0^{s,2} - e_0^{s,1}  \\
 &= \sum _x^\omega {}_n\Delta_x^s \quad \textrm{,}\nonumber 
\end{align}

\noindent where $\omega$ is the highest (open) age group, and $\Delta^s$ is the subgroup-specific ($s$) difference in life expectancy being decomposed, and it is composed of age-specific contributions, ${}_n\Delta_x^s$, which can be calculated following Arriaga's decomposition method following eq \eqref{eq_arriaga1}. I'll use a lifetable radix of 1, meaning $\ell_0 = 1$ to reduce the formula a bit.

\begin{equation}
\label{eq_arriaga1}
 {}_n\Delta_x^s = \begin{cases}\ell_x^{s,1} \cdot \left( \frac{{}_nL_x^{s,2}}{\ell_x^{s,2}} - \frac{{}_nL_x^{s,1}}{\ell_x^{s,1}}\right)+T_{x+n}^{s,2}\cdot\left(\frac{\ell_x^{s,1}}{\ell_x^{s,2}} - \frac{\ell_{x+n}^{s,1}}{\ell_{x+n}^{s,2}}\right)  & \text{$ \forall x < \omega$}  \\
  \ell_\omega ^{s,1} \cdot (e_\omega^{s,2} -e_\omega^{s,1} ) & \text{$\forall x = \omega $}
  \end{cases}
\end{equation}

Equation \eqref{eq_arriaga1} is the first pass of our symmetrical decomposition, where we move _forward_ in time from $t=1$ to $t=2$, so please permit . It

@@

@@

### 2024-01-24 manuscript/manuscript.tex
updates

%% \theoremstyle{thmstylethree}%
\theoremstyle{definition}
\newtheorem{definition}{Definition}%

@@

\abstract{\textbf{Background}: An Arriaga (1984) decomposition allows us
to decompose differences in life expectancy into contributions due to
mortality rate differences in each age. A Kitawaga (1955) decomposition
allows us to decompose differences in a weighted mean into effects from
differences in structure and differences in the weighted value.

@@

Arriaga decompositions can be implemented in a number of relevant ways,
per Riffe et. al.~(2024), from which we here consider only symmetrical
Arriaga decompositions. This implies following the original formulas
twice, once in each direction, and averaging the sign-adjusted results.
For this, we use the following lifetable columns:

@@

We also use the superscript \(s\) to index subpopulations comprising the
total population, and the superscript \(t\) to index time points, or
some other way of differentiating the total population. The index
\(e^{s,t}_0\) reads as ``life expectancy at birth for subpopulation
\(s\) at time \(t\)'', e.g.~\(e^{low, 1990}_0\) could be the life
expectancy in 1990 of a low education group.

Then one can write Arriaga's standard decomposition method like as per
eq\textasciitilde{}\eqref{eq_arriaga1}. To keep things simple, we assume
a lifetable radix of 1, meaning \(\ell_0 = 1\).

\[
\begin{aligned}
{}_n\Delta_x &= e_0^{s,2} - e_0^{s,1} \\
 &= \ell_x^{s,1} \cdot \left( \frac{{}_nL_x^{s,2}}{\ell_x^{s,2}} - \frac{{}_nL_x^{s,1}}{\ell_x^{s,1}}\right)+T_{x+n}^{s,2}\cdot\left(\frac{\ell_x^{s,1}}{\ell_x^{s,2}} - \frac{\ell_{x+n}^{s,1}}{\ell_{x+n}^{s,2}}\right)
\end{aligned}
\]

@@

\begin{flushleft}
Editorial Policies for:

\noindent Springer journals and proceedings:
\url{https://www.springer.com/gp/editorial-policies}

\noindent Nature Portfolio journals:
\url{https://www.nature.com/nature-research/editorial-policies}

\noindent \textit{Scientific Reports}:
\url{https://www.nature.com/srep/journal-policies/editorial-policies}

\noindent BMC journals:
\url{https://www.biomedcentral.com/getpublished/editorial-policies}

\end{flushleft}

>>>>

%% \theoremstyle{thmstylethree}%
\theoremstyle{definition}
\newtheorem{definition}{Definition}%
\usepackage{amsmath}

@@

\abstract{\textbf{Background}: An Arriaga (1984) decomposition allows us
to decompose differences in life expectancy into contributions due to
mortality rate differences in each age. A Kitawaga (1955) decomposition
allows us to decompose differences in a weighted mean into an effect
from differences in structure and effects from differences in each
element of the weighted value.

@@

Arriaga decompositions can be implemented in a number of relevant ways,
per Riffe et. al.~(2024), from which we here consider only
\emph{symmetrical} Arriaga decomposition. This implies following the
original formulas twice, once in each direction, and averaging the
sign-adjusted results. For this, we use the following lifetable columns:

@@

We also use the superscript \(s\) to index subpopulations comprising the
total population, and the superscript \(t\) to index time points, or
some other way of differentiating the total population. The index
\(e^{s,t}_0\) reads as ``life expectancy at birth for subpopulation
\(s\) at time \(t\)'', e.g.~\(e^{low, 1990}_0\) could be the life
expectancy in 1990 of a low education group. The first stage of our
exercise is to decompose subgroup-specific differences in life
expectancy (hence the \(^s\) on everything):

\begin{align}
{}\Delta^s &= e_0^{s,2} - e_0^{s,1}  \\
 &= \sum _x^\omega {}_n\Delta_x^s \quad \textrm{,}\nonumber 
\end{align}

\noindent where \(\omega\) is the highest (open) age group, and
\(\Delta^s\) is the subgroup-specific (\(s\)) difference in life
expectancy being decomposed, and it is composed of age-specific
contributions, \({}_n\Delta_x^s\), which can be calculated following
Arriaga's decomposition method following eq \eqref{eq_arriaga1}. I'll
use a lifetable radix of 1, meaning \(\ell_0 = 1\) to reduce the formula
a bit.

\begin{equation}
\label{eq_arriaga1}
 {}_n\Delta_x^s = \begin{cases}\ell_x^{s,1} \cdot \left( \frac{{}_nL_x^{s,2}}{\ell_x^{s,2}} - \frac{{}_nL_x^{s,1}}{\ell_x^{s,1}}\right)+T_{x+n}^{s,2}\cdot\left(\frac{\ell_x^{s,1}}{\ell_x^{s,2}} - \frac{\ell_{x+n}^{s,1}}{\ell_{x+n}^{s,2}}\right)  & \text{$ \forall x < \omega$}  \\
  \ell_\omega ^{s,1} \cdot (e_\omega^{s,2} -e_\omega^{s,1} ) & \text{$\forall x = \omega $}
  \end{cases}
\end{equation}

Equation \eqref{eq_arriaga1} is the first pass of our symmetrical
decomposition, where we move \emph{forward} in time from \(t=1\) to
\(t=2\), so please permit . It

@@
